---
title: "Multilevel Survival Models"
output:
  bookdown::html_document2: default
  bookdown::pdf_document2: default
bibliography: multilevel_survival.bib
---

```{r set-here, echo=FALSE, include=FALSE}
here::i_am("multilevel-survival/multilevel_survival.Rmd")
```

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  include = TRUE
  )
```

```{r load-packages}
library(survival)
```

# Notation and set-up

The notation basically follows that used in
[@Therneau2000]. For a subject $i$, $T_i^*$ will denote the random variable
corresponding
to the time to the event of interest, and $C_i^*$ will be the censoring time.
The observed follow-up time is $T_i = \min(T_i^*, C_i^*)$, and the indicator
for whether the event was observed for subject $i$ is
$\delta_i = I(T_i^* \leq C_i^*)$, where $I(x) = 1$ if the condition
$x$ is true, and $0$ otherwise. The event time $T_i^*$ is taken to
have probability density function $f_i(t)$ and survival function
$S_i(t) = \mathrm{Pr}(T_i^* > t)$. The corresponding hazard function
is denoted
\begin{equation}
\lambda_i(t) = \lim_{\Delta t \downarrow 0}
\frac{\mathrm{Pr}(t \leq T_i^* < t + \Delta t | T^*_i \geq t)}{\Delta t} =
\frac{f_i(t)}{S_i(t)} ,
\end{equation}
and the cumulative hazard is
\begin{equation}
\Lambda_i(t) = \int_0^t \lambda_i(s) ds =-\log S_i(t).
\end{equation}

When the counting process formulation is used, $N_i(t)$ will
denote the number of observed events in the closed interval $[0, t]$,
and $Y_i(t)$ will be an indicator for whether subject $i$ is observed
and at risk at time $t$. For right-censored survival data with a single event
per subject,
\begin{equation}
\begin{aligned}
N_i(t) &= I(\{ T_i \leq t , \delta_i = 1 \}) , \\
Y_i(t) &= I(T_i \geq t) .
\end{aligned}
\end{equation}
For $n$ subjects being observed, the size of the risk set and the total
number of events are denoted $\overline{Y}(t) = \sum_{i=1}^n Y_i(t)$ and
$\overline{N}(t) = \sum_{i=1}^n N_i(t)$.

Any covariates measured for the $i^{\text{th}}$ subject will be collected
into a vector $\mathbf{x}_i(t)$. In general, this may include both
time-independent and time-dependent variables, although for simplicity
often the explicit time-dependence is not shown.
A proportional hazards model supposes that the hazard function
can be modelled in terms of these covariates as
\begin{equation} \label{eq:ph-model}
\lambda_i(t | \mathbf{x}_i(t)) = \lambda_0(t)
\exp[ \mathbf{x}^\prime_i(t) \beta ] ,
\end{equation}
where $\beta$ is a vector of regression coefficients and $\lambda_0(t)$
is a baseline hazard. The baseline hazard may have a parametric model
specified for it, or it may be left unspecified as in the Cox
model [@Cox1972].

If the form of the baseline hazard were known, and assuming
independent and non-informative
censoring, the likelihood function
for the observed times for $n$ subjects could be evaluated as
\begin{equation}
L(\lambda_0(t), \beta) = \prod_{i=1}^n f_i(t|\mathbf{x}_i)^{\delta_i} S_i(t |\mathbf{x}_i)^{1 - \delta_i} =
\prod_{i=1}^n [\lambda_i(t|\mathbf{x}_i) S_i(t|\mathbf{x}_i) ]^{\delta_i}
S_i(t|\mathbf{x}_i)^{1 - \delta_i} =
\prod_{i=1}^n \lambda_i(t|\mathbf{x}_i)^{\delta_i} S_i(t|\mathbf{x}_i) ,
\end{equation}
resulting in the log-likelihood
\begin{equation}
\ell(\lambda_0(t), \beta) = \sum_{i=1}^n \left \{ \delta_i [ \log \lambda_0(t) + \mathbf{x}^\prime_i(t) \beta ] -
\int_0^t \lambda_0(s) \exp [ \mathbf{x}_i^\prime(s) \beta ] ds \right \}.
\end{equation}
In the special case where all covariates are time-independent,
\begin{equation}
\ell(\lambda_0(t), \beta) = \sum_{i=1}^n \left \{ \delta_i [ \log \lambda_0(t) + \mathbf{x}^\prime_i \beta ] -
\Lambda_0(t) \exp [ \mathbf{x}_i^\prime \beta ] \right \}.
\end{equation}
When $\lambda_0(t)$ is left completely unspecified, we rewrite the
likelihood as
\begin{equation}
\begin{aligned}
\ell(\lambda_0(t), \beta) &= \sum_{i=1}^n \left \{ \delta_i \left [ 
\mathbf{x}^\prime_i \beta - \log \sum_{j \in R_i} \exp[
\mathbf{x}^\prime_j \beta ] \right ] \right \} \\
\quad & {} + \sum_{i=1}^n \left \{ \delta_i \left [
\log \lambda_0(t) + \log \sum_{j \in R_i} \exp[
\mathbf{x}^\prime_j \beta ] \right ] -
\Lambda_0(t) \exp [ \mathbf{x}_i^\prime \beta ] \right \},
\end{aligned}
\end{equation}
where $R_i$ is the risk set at time $t_i$.

# Multivariate Frailty Model

The notation in this section follows [@Ripatti2000].

```{r session-info}
sessionInfo()
```
