---
title: "Rates and Standardisation"
output:
  html: default
bibliography: rates_and_standardisation.bib
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
    echo = TRUE,
    include = TRUE
)
```

```{r}
#| label: load-packages
library(Epi)
library(knitr)
library(kableExtra)
library(ggplot2)
```

Standardisation allows sets of stratum-specific measures,
such as incidence rates or proportions, to be summarised
in such a way that the effects of confounding factors are
adjusted for.

To take the typical example, disease incidence rates frequently
vary with both age and calendar time. To compare the incidence
of disease in two or more groups, one might imagine using as
a summary measure the crude incidence rate observed for each
group. If the groups differ substantially in their age
distributions, however, then such a comparison is likely to be
misleading (@tbl-crude-rate-exmpl). When appropriate,
age-standardisation allows for the confounding effects of
age to be addressed in such cases.

```{r}
#| echo: false
#| label: tbl-crude-rate-exmpl
#| tbl-cap: "A simple example of how comparisons of crude measures may be confounded by differences in age structures. In this case, the two populations have identical age-specific rates, but the overall crude incidence rates differ due to the dramatically different age structures."
exmpl_data <- data.frame(
    r = c(0.01, 0.01, 0.02, 0.05, 0.1),
    agegroup = c("0-19", "20-39", "40-59", "60-79", "80+"),
    n1 = c(500, 400, 300, 200, 100),
    n2 = c(100, 200, 300, 400, 500)
)

exmpl_data$d1 <- exmpl_data$r * exmpl_data$n1
exmpl_data$r1 <- 100 * exmpl_data$d1 / exmpl_data$n1
exmpl_data$d2 <- exmpl_data$r * exmpl_data$n2
exmpl_data$r2 <- 100 * exmpl_data$d2 / exmpl_data$n2

exmpl_data <- exmpl_data[, c("agegroup", "n1", "d1", "r1",
                              "n2", "d2", "r2")]

totals <- data.frame(
    agegroup = c("Total"),
    n1 = sum(exmpl_data$n1),
    d1 = sum(exmpl_data$d1),
    n2 = sum(exmpl_data$n2),
    d2 = sum(exmpl_data$d2)
)
totals$r1 <- 100 * totals$d1 / totals$n1
totals$r2 <- 100 * totals$d2 / totals$n2

exmpl_data <- rbind(exmpl_data, totals)

exmpl_data$ratio <- exmpl_data$r2 / exmpl_data$r1

colnames(exmpl_data) <- c(
    "Age Group",
    "Observed Person-years", "Number of Cases",
    "Incidence Rate per 100 Person-years",
    "Observed Person-years", "Number of Cases",
    "Incidence Rate per 100 Person-years", "Rate Ratio")

kable(exmpl_data, digits = 2) %>%
  row_spec(nrow(exmpl_data), bold = TRUE) %>%
  add_header_above(c(" " = 1, "Population A" = 3,
                     "Population B" = 3, " " = 1))
```

These notes will mostly follow @Breslow1987.
For a given cohort or population, let $d_j$ be the
observed number of events in the $j^{\text{th}}$ stratum
($j = 1, \dots, J$) and let $n_j$ be the number of persons
initially at risk.

When the population is observed over a given time interval,
and assuming that during the interval there is no entry of
additional persons (e.g., due to immigration), nor do people leave
without their event status being determined (e.g., due to
emigration), one measure of incidence that may be considered
is the incidence proportion, i.e., the proportion of people
in the initial population at risk who experience the
event during the
interval. An estimate for the incidence proportion within the
$j^{\text{th}}$ stratum is
\begin{equation}
\hat{p}_j = \frac{d_j}{n_j} .
\end{equation}
The incidence proportion is a measure of the average risk of
the event occurring for persons within the stratum.

The incidence proportion does not make use the amount of time
each person under study is followed for before experiencing
(or not experiencing) the event of interest. A total of
10 events occurring in a cohort of 100 persons equates to an
incidence proportion of 0.1, whether those events occurred
over the space of 1 year or 10 years. The rate at which
events occur in these two situations
is likely to be quite different: in the former case, we would
expect the average time to experiencing the event to be
much shorter. The incidence rate, which accounts for the total
time each person in the population is observed to be at risk,
is obtained by taking the denominator to be the total
person-time at risk.
Let $\tau_j$ be the total observed person-time in the
$j^{\text{th}}$ stratum.
Then the incidence rate for the stratum is estimated as
\begin{equation}
\hat{\lambda}_j = \frac{d_j}{\tau_j} .
\end{equation}

When stratum membership is time-invariant, calculation of
$\tau_j$ is fairly straightforward, by simply summing the
total time at risk for all individuals in the stratum.
Some care is needed, however, in determining the time at
risk for each individual. In particular, only follow-up intervals
during which the individual is under observation and at risk
of experiencing the event should be counted.
Periods where the individual could not possibly have
experienced the event (i.e., was not at risk) should not be
counted. The typical example occurs when entry into a cohort
is conditional on the individual reaching a particular
milestone (years of exposure, years on treatment, etc.)
without experiencing the event of interest. In this case,
the time prior to reaching the milestone should not
contribute to the person-time denominator. Time after a
person ceases to be at risk should also not be counted.
For example, if the event of interest is defined as the
first occurrence of some potentially recurrent outcome,
then each person in the population would only contribute
time at risk up until first experiencing the event.
For individuals who are lost to follow-up and
who hence cannot have their event status ascertained,
contributions to time at risk should cease
at the last time their status is known.

The calculation is more involved if stratum membership is
time-varying, for example when strata are defined by age
and calendar period, or when needed for the purposes of
standardisation. In this case it is necessary to
break up the follow-up period into the intervals where the
person is in each stratum, and allocate the corresponding
amount of person-time to the denominator for the relevant
stratum. Consider the example given in @Hill1972 of a
person who is born on 16 July 1928, enters a study on
5 January 1952, and exits on 30 June 1962. The person's
contribution to the person-years denominator in each
5-year age group 0-4, 5-9, etc. is calculated as follows:
```{r}
#| label: hill-pyears-exmpl
# Example calculation of person-years as in Hill 1972
date_birth <- as.Date("1928-07-16")
date_entry <- as.Date("1952-01-05")
date_exit <- as.Date("1962-05-30")

# For convenience, convert dates to fractional years
# (assuming every year is 365.25 days long)
year_birth <- as.numeric(date_birth) / 365.25 + 1970
year_entry <- as.numeric(date_entry) / 365.25 + 1970
year_exit <- as.numeric(date_exit) / 365.25 + 1970

# Calculate age at entry and exit
age_entry <- year_entry - year_birth
age_exit <- year_exit - year_birth

# Define five-year age groups
age_breaks <- seq(5, 85, by = 5)

# For each interval, determine the time of entry into
# the interval on the age time-scale, which is either
# the person's age at entry or the left end of the interval
# (whichever is the larger)
entry_time <- pmax(age_entry, c(-Inf, age_breaks))

# Similarly, work out the time of exit from the interval
# on the age time-scale - either the person's age at exit
# or the right end of the interval (whichever is the smallest)
exit_time <- pmin(age_exit, c(age_breaks, Inf))

# Calculate the duration spent in each interval,
# setting person-time to zero for intervals where the
# person does not contribute any time
pers_years <- exit_time - entry_time
mask <- exit_time < entry_time
pers_years[mask] <- 0

# For neater printing
pers_years_summ <- data.frame(
    age_start = c(0, age_breaks),
    age_end = c(age_breaks, Inf),
    person_years = pers_years
)

pers_years_summ[pers_years_summ$person_years > 0,]
```

Note that, because the exact dates of birth, entry and
exit are available, an exact calculation of the person-years
contributed to each age group is possible. When the exact dates
are not known, it is necessary to approximate the person-time
spent in each interval. @Breslow1987 give the example scheme
of taking 0.5 person-years for the year of entry and exit, and
1 full person-year for each intervening year, or including
a single 0.25 person-year contribution if the year of entry
and exit coincide.

When stratum membership varies across multiple time
dimensions (e.g., current age and calendar year, or current
age and time on treatment), the above calculation needs to
be extended to compute the amount of time spent at risk
by each person within each hypercube defined by the
breaks used to discretise each time scale. Sticking with
the example in @Hill1972, one must calculate the person-years
at risk by 5-year age group and calendar period. In
@fig-hill-exmpl-graph, this corresponds to the length of the
line within each grid cell.
```{r}
#| echo: false
#| label: fig-hill-exmpl-graph
#| fig-cap: Example of distributing person-years at risk across age and calendar period intervals
hill_exmpl <- data.frame(
    age = c(age_entry, age_exit),
    period = c(date_entry, date_exit)
)

p <- ggplot(hill_exmpl, aes(x = period, y = age)) +
    geom_line() +
    scale_y_continuous(breaks = seq(20, 40, by = 5),
                       limits = c(20, 40),
                       minor_breaks = NULL) +
    scale_x_date(breaks = c(as.Date("1951-01-01"),
                            as.Date("1956-01-01"),
                            as.Date("1961-01-01"),
                            as.Date("1966-01-01")),
                 limits = c(as.Date("1951-01-01"),
                            as.Date("1966-01-01")),
                 minor_breaks = NULL) +
    labs(
        x = "Date",
        y = "Age"
    ) +
    theme_bw()

p
```

In R, the necessary calculations can be done using (for example)
the Epi package:
```{r}
#| label: hill-exmpl-epi
# Define the date of birth, date of entry into the study,
# and date of exit for our example person
hill_exmpl <- data.frame(
    date_birth = cal.yr(as.Date("1928-07-16")),
    date_entry = cal.yr(as.Date("1952-01-05")),
    date_exit = cal.yr(as.Date("1962-05-30")),
    status = factor(c(FALSE), labels = c("Censored"))
)

# Create a Lexis object, specifying each of the relevant
# time scales (per = calendar time, age = age, tfE = time
# from entry into the study)
hill_lexis <- Lexis(
    entry = list(
        per = date_entry,
        age = date_entry - date_birth,
        tfE = 0
    ),
    exit = list(per = date_exit),
    exit.status = status,
    data = hill_exmpl
)

# Split the follow-up time for the example person by
# 5-year age and calendar interval, either in two steps ...
hill_split1 <- splitLexis(
    hill_lexis, "age", breaks = seq(0, 100, by = 5))
hill_split1 <- splitLexis(
    hill_split1, "per", breaks = seq(1951, 1966, by = 5)
)

print(hill_split1)

# ... or one
hill_split2 <- popEpi::splitMulti(
    hill_lexis,
    age = seq(0, 100, by = 5),
    per = seq(1951, 1966, by = 5),
    drop = FALSE
)

print(hill_split2)
```

## Direct standardisation





## Session details

```{r}
#| label: session-info
sessionInfo()
```