---
title: "Rates and Standardisation"
output:
  html: default
bibliography: rates_and_standardisation.bib
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
    echo = TRUE,
    include = TRUE
)
```

```{r}
#| label: load-packages
library(Epi)
library(knitr)
library(kableExtra)
```

Standardisation allows sets of stratum-specific measures,
such as incidence rates or proportions, to be summarised
in such a way that the effects of confounding factors are
adjusted for.

To take the typical example, disease incidence rates frequently
vary with both age and calendar time. To compare the incidence
of disease in two or more groups, one might imagine using as
a summary measure the crude incidence rate observed for each
group. If the groups differ substantially in their age
distributions, however, then such a comparison is likely to be
misleading (@tbl-crude-rate-exmpl). When appropriate,
age-standardisation allows for the confounding effects of
age to be addressed in such cases.

```{r}
#| echo: false
#| label: tbl-crude-rate-exmpl
#| tbl-cap: "A simple example of how comparisons of crude measures may be confounded by differences in age structures. In this case, the two populations have identical age-specific rates, but the overall crude incidence rates differ due to the dramatically different age structures."
exmpl_data <- data.frame(
    r = c(0.01, 0.01, 0.02, 0.05, 0.1),
    agegroup = c("0-19", "20-39", "40-59", "60-79", "80+"),
    n1 = c(500, 400, 300, 200, 100),
    n2 = c(100, 200, 300, 400, 500)
)

exmpl_data$d1 <- exmpl_data$r * exmpl_data$n1
exmpl_data$r1 <- 100 * exmpl_data$d1 / exmpl_data$n1
exmpl_data$d2 <- exmpl_data$r * exmpl_data$n2
exmpl_data$r2 <- 100 * exmpl_data$d2 / exmpl_data$n2

exmpl_data <- exmpl_data[, c("agegroup", "n1", "d1", "r1",
                              "n2", "d2", "r2")]

totals <- data.frame(
    agegroup = c("Total"),
    n1 = sum(exmpl_data$n1),
    d1 = sum(exmpl_data$d1),
    n2 = sum(exmpl_data$n2),
    d2 = sum(exmpl_data$d2)
)
totals$r1 <- 100 * totals$d1 / totals$n1
totals$r2 <- 100 * totals$d2 / totals$n2

exmpl_data <- rbind(exmpl_data, totals)

exmpl_data$ratio <- exmpl_data$r2 / exmpl_data$r1

colnames(exmpl_data) <- c(
    "Age Group",
    "Observed Person-years", "Number of Cases",
    "Incidence Rate per 100 Person-years",
    "Observed Person-years", "Number of Cases",
    "Incidence Rate per 100 Person-years", "Rate Ratio")

kable(exmpl_data, digits = 2) %>%
  row_spec(nrow(exmpl_data), bold = TRUE) %>%
  add_header_above(c(" " = 1, "Population A" = 3,
                     "Population B" = 3, " " = 1))
```

For a given cohort or population, let $d_j$ be the
observed number of events in the $j^{\text{th}}$ stratum
($j = 1, \dots, J$) and let $n_j$ be the number of persons
initially at risk.

When the population is observed over a given time interval,
and assuming that during the interval there is no entry of
additional persons (e.g., due to immigration), nor do people leave
without their event status being determined (e.g., due to
emigration), one measure of incidence that may be considered
is the incidence proportion, i.e., the proportion of people
in the initial population at risk who experience the
event during the
interval. An estimate for the incidence proportion within the
$j^{\text{th}}$ stratum is
\begin{equation}
\hat{p}_j = \frac{d_j}{n_j} .
\end{equation}
The incidence proportion is a measure of the average risk of
the event occurring for persons within the stratum.

The incidence proportion does not make use the amount of time
each person under study is followed for before experiencing
(or not experiencing) the event of interest. A total of
10 events occurring in a cohort of 100 persons equates to an
incidence proportion of 0.1, whether those events occurred
over the space of 1 year or 10 years. The rate at which
events occur in these two situations
is likely to be quite different: in the former case, we would
expect the average time to experiencing the event to be
much shorter. 

Let $\tau_j$ be the total observed person-time in the
$j^{\text{th}}$ stratum.
Then the incidence rate for the stratum is estimated as
\begin{equation}
\hat{\lambda}_j = \frac{d_j}{\tau_j} .
\end{equation}



## Direct standardisation



[@Breslow1987]

## Session details

```{r}
#| label: session-info
sessionInfo()
```