---
title: "Differential Expression Analysis"
output:
  bookdown::html_document2: default
bibliography: differential_expression.bib
---


```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = TRUE,
  include = TRUE
  )
```

# Linear Models

Approach underlying, e.g., limma [@Ritchie2015]. Formulated
initially in the context of analysis of microarray data,
e.g. [@Smyth2004]. For the $g^{\text{th}}$ gene, let
$y_{g1}, \dots, y_{gn}$ be the measured response values. For
microarray data, for example, $y_{gi}$ is usually either
the log-ratio $\log_2 R_{gi} - \log_2 G_{gi}$ (for two-colour
arrays) or log-intensity (for one colour arrays) measured
for gene $g$ in array $i$.

After normalising to remove unwanted technical variation, the
response is assumed to be described by the linear model
\begin{equation}
\mathrm{E}[\mathbf{y}_g] = X \mathbf{\alpha}_g ,
\end{equation}
where $\mathbf{y}_g^\prime = (y_{g1},\dots,y_{gn})$, $X$ is 
a suitable design matrix, and $\mathbf{\alpha}_g$ is the
coefficient vector. The variance is allowed to vary across
genes, such that
\begin{equation}
\mathrm{Var}[\mathbf{y}_g] = W_g \sigma_g^2 ,
\end{equation}
where $W_g$ is a (known) non-negative matrix of weights.

Biological contrasts of interest are encoded as
\begin{equation}
\mathbf{\beta}_g = C^\prime \mathbf{\alpha}_g ,
\end{equation}
for suitable $C$. The general problem in a differential
expression analysis is to assess whether particular contrasts
within this set are zero.


```{r}
#| label: session-info
sessionInfo()
```
